---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ipdcstorm

<!-- badges: start -->
<!-- badges: end -->

`ipdcstorm` is an R package for stochastic tropical-cyclone wind hazard
modelling at point locations in the North Atlantic basin. It was developed
to support climate stress testing of supply-chain infrastructure on the
Dutch Caribbean islands (St. Martin, Saba, Statia), but it works for any
site in the basin.

The package combines NOAA's IBTrACS historical track database with a
parametric wind model and a stochastic event generator to produce long
synthetic records of daily site-level wind exposure under both stationary
and climate-change conditions.

## What it does

The workflow has three stages:

1. **Historical calibration.** Reads IBTrACS tracks, computes site-level
   winds with a Holland (1980) parametric profile (with forward-motion
   asymmetry and directional radii), and estimates per-site annual storm
   rates and Gamma overdispersion by severity class.
2. **Stochastic simulation.** Draws synthetic seasons from a
   Gamma–Poisson–Binomial process that reproduces observed burstiness and
   the tropical-storm/hurricane split. Climate change enters through SST
   forcing — affecting both annual rates and the hurricane fraction —
   with KNMI'23 and IPCC AR6 scenarios built in.
3. **Temporal downscaling.** Resamples historical events from a
   severity-stratified library, places them on the calendar via empirical
   day-of-year distributions, and spreads each event over its duration
   with a smooth wind envelope to produce a daily wind (kt) time series
   per location.

It also ships validation utilities (rates, return levels, wind-field
diagnostics, QQ/CDF/bias plots) and a hindcast suite for checking the
calibrated model against the historical record.

## Installation

Install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("tanerumit/ipdcstorm")
```

## Quick start

A minimal end-to-end run for three Dutch Caribbean sites using the
packaged IBTrACS demo subset:

```{r, eval = FALSE}
library(ipdcstorm)

targets <- tibble::tribble(
  ~name,       ~lat,     ~lon,
  "St_Martin", 18.0708, -63.0501,
  "Saba",      17.6350, -63.2300,
  "Statia",    17.4890, -62.9740
)

cfg <- make_hazard_cfg(
  data_path        = system.file("extdata", "ibtracs_demo.csv",
                                 package = "ipdcstorm"),
  simulation_years = 1000,
  climate          = make_climate_cfg(scenario = "stationary")
)

out <- run_hazard_model(cfg, targets = targets, seed = 42)

daily <- generate_daily_hazard_impact_spatial(
  out         = out,
  location    = targets$name,
  sim_years   = seq_len(1000),
  year0       = 2000L,
  damage      = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario    = "baseline",
  seed        = 42
)
```

`daily` is a named list with one tibble per location. Each per-location table
contains `sim_year`, `date`, `wind_kt`, `surge_m`, `event_id`,
`pressure_hpa`, `pressure_deficit_hpa`, `rmw_km`, `damage_intensity`,
`damage_rate`, and `cum_damage`.

For production runs, replace `data_path` with the full IBTrACS North
Atlantic CSV (`ibtracs.NA.list.v04r01.csv`), available from NOAA NCEI.

## Climate scenarios

Climate forcing is configured through `make_climate_cfg()` and supports:

- **Stationary** baseline (no warming).
- **KNMI'23** scenarios at 2050 and 2100 horizons (`knmi_Ld`, `knmi_Hd`).
- **IPCC AR6** SST scenarios.
- **Direct** `delta_sst` for custom forcing.

Scenarios shift annual storm rates, the hurricane share, and (optionally)
individual event intensities, while preserving common random numbers
between baseline and scenario runs so differences isolate the climate
signal.

## Tutorials

Three Quarto tutorials walk through the full workflow:

- `vignettes/tutorial_1_setup.qmd` — model overview, setup, single-site
  baseline.
- `vignettes/tutorial_2_validation.qmd` — validation suite and
  hindcasting.
- `vignettes/tutorial_3_climate_change.qmd` — multi-site climate-change
  analysis.

Worked example scripts live in `inst/extcode/`, including a
stress-test experiment (`05-stress-test-experiment.R`) that selects
representative high-impact years and replays them under KNMI'23
scenarios.

## Key entry points

| Function | Purpose |
|---|---|
| `make_hazard_cfg()` / `run_hazard_model()` | Configure and run the hazard simulation |
| `make_climate_cfg()` / `resolve_climate_inputs()` | Build a climate scenario |
| `generate_daily_hazard_impact_spatial()` | Daily downscaling for one or more sites |
| `make_validation_cfg()` / `run_validation_suite()` | Validate against the historical record |
| `plot_*()` / `save_hazard_viz_plots()` | Diagnostic and reporting plots |

## License

MIT. See `LICENSE`.

## Issues

Please report bugs at
<https://github.com/tanerumit/ipdcstorm/issues>.
