---
title: "Tutorial 1 — Introduction, Setup & Single-Site Baseline"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

This tutorial walks through the smallest useful `ipdcstorm` workflow: configure a stationary baseline, run the hazard model for one location, and convert annual simulations into a daily wind-impact series. The goal is to give you a reproducible first run that uses the current package API and produces outputs you can inspect immediately.

The workflow has four steps:

1. Load the package and example IBTrACS file.
2. Build a `hazard_cfg` with an embedded climate configuration.
3. Run `run_hazard_model()` for one site.
4. Downscale selected synthetic years with `generate_daily_hazard_impact()`.

## Prepare the Session {#sec-setup}

Start by loading the package and the small example dataset that ships with `ipdcstorm`. The demo file is enough for a quick tutorial render; for analysis, point `data_path` to the full North Atlantic IBTrACS file.

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
library(tibble)

ibtracs_demo <- system.file("extdata", "ibtracs_demo.csv", package = "ipdcstorm")
ibtracs_demo
```

::: callout-important
## Data requirement
The tutorial uses packaged demo data for speed. Production runs should use the North Atlantic IBTrACS archive because annual rates, event counts, and the empirical event library are sensitive to the historical record you provide.
:::

## Build a Baseline Configuration {#sec-configure}

Climate is always carried inside `hazard_cfg`; there is no separate climate-off mode. A stationary baseline is represented explicitly with `make_climate_cfg(scenario = "stationary")`, which resolves to a zero SST shift while preserving the same workflow used for future scenarios.

```{r}
#| label: configure-climate-baseline
climate_baseline <- make_climate_cfg(scenario = "stationary")
climate_baseline
```

Now create the hazard configuration. The main arguments you usually need to think about are the IBTrACS path, the historical calibration window, the synthetic simulation horizon, and the search radius around each target.

```{r}
#| label: configure-hazard-baseline
cfg <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  simulation_years = 200L,
  search_radius_km = 800,
  climate = climate_baseline
)

cfg
```

::: callout-note
## Parameters to tune first
| Parameter | Why it matters |
|---|---|
| `historical_start_year` | Sets the first year used to calibrate annual storm rates. A later start year can reduce older-data inconsistencies but shortens the calibration sample. |
| `simulation_years` | Controls Monte Carlo stability. Use a small value for debugging and a much larger value when estimating rare-event probabilities. |
| `search_radius_km` | Filters track points before site-wind calculation. It is a computational gate, not a physical cutoff for wind impact. |
| `climate` | Encodes whether the run is stationary or climate-adjusted. Even a baseline run should supply an explicit climate configuration. |
:::

## Define the Target Site {#sec-target}

Targets are simple tables with `name`, `lat`, and `lon`. Here we use Saba as a single-site example.

```{r}
#| label: setup-target-saba
targets <- tribble(
  ~name,  ~lat,   ~lon,
  "Saba", 17.635, -63.230
)

targets
```

## Run the Annual Hazard Model {#sec-run}

`run_hazard_model()` reads the IBTrACS data, computes site-level winds for nearby track points, aggregates them into historical events, estimates annual rates, and simulates synthetic yearly counts. Setting a seed keeps the tutorial deterministic.

```{r}
#| label: run-hazard-baseline
out <- run_hazard_model(
  cfg = cfg,
  targets = targets,
  seed = 42,
  verbose = FALSE
)
```

Inspect the object structure first. The main outputs for a first pass are `rates`, `fit`, `sim`, `events`, and `trackpoints`.

```{r}
#| label: inspect-output-names
names(out)
```

The annual rates summarize how often tropical-storm and hurricane-strength events affect the site in the historical calibration period.

```{r}
#| label: inspect-rates-table
out$rates
```

The `fit` table adds the overdispersion estimate `k_hat` and the resolved climate metadata used in the simulation. In a stationary baseline, the climate fields should remain neutral.

```{r}
#| label: inspect-fit-table
out$fit
```

The synthetic annual series is the bridge to downstream risk analysis. Each row is one simulated year with tropical-storm and hurricane counts.

```{r}
#| label: inspect-simulated-years
head(out$sim)
```

Historical events remain available for diagnostics and for building the empirical event library used later in daily downscaling.

```{r}
#| label: inspect-historical-events
head(out$events)
```

## Downscale to Daily Wind and Damage {#sec-daily}

The annual model tells you how many storms occur in each synthetic year, but many downstream models need daily forcing. `generate_daily_hazard_impact()` resamples events from the historical library, assigns event timing within each year, applies a pulse shape, and derives gust wind, surge proxy, and damage variables.

The function returns a named list, even when you request a single location, so extract the site you want from that list.

```{r}
#| label: run-daily-downscaling
daily_list <- generate_daily_hazard_impact(
  out = out,
  location = "Saba",
  sim_years = 1:20,
  year0 = 2025,
  gust_factor = 1.25,
  damage = list(
    method = "powerlaw",
    thr = 34,
    V_ref = 80,
    d_ref = 0.03,
    p = 3,
    d_max = 0.10
  ),
  scenario = "stationary",
  seed = 42
)

daily_saba <- daily_list$Saba
head(daily_saba)
```

This quick summary shows how much hazard activity appears in the selected synthetic years. `tc_days` counts any day with non-zero wind from a sampled tropical cyclone, while the threshold counts focus on tropical-storm and hurricane conditions at the site.

```{r}
#| label: summarise-daily-series
daily_saba |>
  summarise(
    total_days = n(),
    tc_days = sum(wind_kt > 0, na.rm = TRUE),
    ts_days = sum(wind_kt >= 34, na.rm = TRUE),
    hur_days = sum(wind_kt >= 64, na.rm = TRUE),
    mean_annual_damage = sum(damage_rate, na.rm = TRUE) / n_distinct(sim_year)
  )
```

## How Climate Enters Future Runs {#sec-climate}

Once the stationary baseline is working, you can swap in a different climate configuration without changing the rest of the workflow. The package supports two equivalent patterns:

- Direct mode: set `delta_sst` explicitly for a transparent sensitivity test.
- Scenario helper mode: set `scenario` plus `target_year` to derive the corresponding SST shift and keep provenance metadata.

```{r}
#| label: configure-climate-futures
climate_delta <- make_climate_cfg(delta_sst = 0.8)
climate_ssp585 <- make_climate_cfg(
  scenario = "ssp585",
  target_year = 2050
)

climate_delta
climate_ssp585
```

Both approaches resolve to a scalar SST shift before the hazard simulation runs. That means the modeling logic stays consistent across baseline, stress-test, and scenario-based runs; what changes is the resolved climate input and its metadata.

## Summary {#sec-summary}

You now have a complete first run with the current `ipdcstorm` interface: an explicit baseline climate configuration, a single-site hazard simulation, and a daily hazard-impact series for selected synthetic years. The next practical steps are to increase `simulation_years`, add more target locations, or replace the demo file with the full IBTrACS record for analysis-grade runs.
