---
title: "Tutorial 1 - Introduction, Setup, and a First Hazard Run"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

This tutorial shows the simplest end-to-end workflow for the hazard model. You will set up one target location, run a stationary baseline simulation, and then see how to switch to climate stress tests using the current climate API.

## What This Tutorial Covers

- Build a `hazard_cfg` with embedded climate settings.
- Run `run_hazard_model()` for a single site with the current argument names.
- Use the three supported climate patterns: stationary, direct `delta_sst`, and scenario plus `target_year`.
- Convert annual simulated counts into a daily hazard-impact series.

Estimated time: 15-20 minutes.

## Overview {#sec-overview}

The model workflow has three practical steps:

1. Create a hazard configuration with IBTrACS input data, simulation settings, and an embedded climate configuration.
2. Run the annual hazard model to calibrate historical rates and simulate synthetic storm counts.
3. Downscale selected simulated years to daily wind and impact time series.

The climate workflow is now intentionally simple. Every run carries a climate configuration inside `hazard_cfg`. A stationary baseline uses `make_climate_cfg(scenario = "stationary")`; a hypothetical stress test uses `make_climate_cfg(delta_sst = ...)`; and a scenario-based run uses `make_climate_cfg(scenario = "ssp585", target_year = ...)`.

## Prepare the Session {#sec-prerequisites}

Load the package and a few helper packages used in the examples.

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
library(tibble)
```

This tutorial uses the small built-in IBTrACS example file so the code runs quickly.

```{r}
#| label: setup-data
ibtracs_demo <- system.file("extdata", "ibtracs_demo.csv", package = "ipdcstorm")
ibtracs_demo
```

Define one target location. A target table must contain `name`, `lat`, and `lon`.

```{r}
#| label: setup-target
targets <- tribble(
  ~name,   ~lat,   ~lon,
  "Saba",  17.635, -63.230
)
```

## Configure a Baseline Run {#sec-configure}

Start with a stationary baseline. In the current API, climate is always embedded in `make_hazard_cfg()`.

```{r}
#| label: configure-baseline-climate
climate_baseline <- make_climate_cfg(scenario = "stationary")
climate_baseline
```

Now create the hazard configuration. The main user-facing arguments are `data_path`, `historical_start_year`, `simulation_years`, `search_radius_km`, and `climate`.

```{r}
#| label: configure-baseline-hazard
cfg_baseline <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  simulation_years = 200L,
  search_radius_km = 800,
  climate = climate_baseline
)

cfg_baseline
```

These settings mean:

| Parameter | Description |
|---|---|
| `data_path` | Path to the IBTrACS CSV used for calibration. |
| `historical_start_year` | First historical year included in the rate calibration. |
| `simulation_years` | Number of synthetic years to generate. Use a small value for tutorials and a larger value for analysis. |
| `search_radius_km` | Distance filter used to retain track points near the target location. |
| `climate` | Climate configuration resolved inside the run. |

## Run the Baseline Hazard Model {#sec-run}

Run the model with the hazard config and target table. The climate settings are already stored in `cfg_baseline`, so you no longer pass a separate climate object to `run_hazard_model()`.

```{r}
#| label: run-baseline
out_baseline <- run_hazard_model(
  cfg = cfg_baseline,
  targets = targets,
  seed = 42,
  verbose = FALSE
)
```

The returned object contains the historical calibration products and the synthetic annual simulations.

```{r}
#| label: inspect-run-objects
names(out_baseline)
```

Inspect the calibrated annual rates.

```{r}
#| label: inspect-rates
out_baseline$rates
```

Inspect the fitted annual-count parameters. In a stationary baseline, the resolved climate metadata should show `delta_sst = 0`.

```{r}
#| label: inspect-fit
out_baseline$fit
out_baseline$run_metadata$climate
```

Inspect the first few simulated years. This is the main output used for downstream summaries and downscaling.

```{r}
#| label: inspect-sim
head(out_baseline$sim)
```

## Choose a Climate Workflow {#sec-climate}

The current climate interface supports two kinds of future runs.

Direct `delta_sst` is the simplest way to stress-test the model. Here `delta_sst` is an additive sea-surface-temperature anomaly in degrees C relative to the baseline climate. It is not a direct storm-count multiplier. Count effects are applied downstream through the model's rate-scaling logic, while intensity effects are handled through the resolved climate scalars inside the annual simulation.

```{r}
#| label: configure-direct-delta
climate_delta <- make_climate_cfg(delta_sst = 0.8)
climate_delta
```

Scenario helper mode is for time-slice runs tied to a named pathway. The scenario and `target_year` are used only to resolve a canonical `delta_sst`, which is then passed through the same downstream logic as the direct mode.

```{r}
#| label: configure-scenario-helper
climate_ssp585 <- make_climate_cfg(
  scenario = "ssp585",
  target_year = 2050
)
climate_ssp585
```

The helper is useful when you want provenance such as "SSP5-8.5 in 2050", while direct `delta_sst` is better for transparent sensitivity testing.

Run short examples of both future patterns by embedding the climate config in a new hazard config.

```{r}
#| label: run-future-examples
cfg_delta <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  simulation_years = 50L,
  search_radius_km = 800,
  climate = climate_delta
)

cfg_ssp585 <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  simulation_years = 50L,
  search_radius_km = 800,
  climate = climate_ssp585
)

out_delta <- run_hazard_model(
  cfg = cfg_delta,
  targets = targets,
  seed = 42,
  verbose = FALSE
)

out_ssp585 <- run_hazard_model(
  cfg = cfg_ssp585,
  targets = targets,
  seed = 42,
  verbose = FALSE
)
```

Compare the resolved climate metadata. Notice that both future workflows end up with a resolved `delta_sst`, but only the scenario-helper run also carries scenario and target-year provenance.

```{r}
#| label: compare-climate-metadata
climate_summary_row <- function(x, example) {
  tibble(
    example = example,
    input_mode = x$input_mode,
    scenario = x$scenario,
    target_year = if (is.null(x$target_year)) NA_real_ else x$target_year,
    delta_sst = x$delta_sst,
    rate_scale = x$rate_scale,
    gamma = x$gamma
  )
}

bind_rows(
  climate_summary_row(out_baseline$run_metadata$climate, "baseline"),
  climate_summary_row(out_delta$run_metadata$climate, "direct_delta"),
  climate_summary_row(out_ssp585$run_metadata$climate, "scenario_helper")
)
```

## Downscale to Daily Hazard Impacts {#sec-daily}

The annual simulation gives counts per synthetic year. To create a daily hazard series, use `generate_daily_hazard_impact()`. The current function returns a named list, so for one location you extract the location table from that list.

```{r}
#| label: run-daily-downscaling
daily_list <- generate_daily_hazard_impact(
  out = out_baseline,
  location = "Saba",
  sim_years = 1:20,
  year0 = 2025,
  gust_factor = 1.25,
  damage = list(
    method = "powerlaw",
    thr = 34,
    V_ref = 80,
    d_ref = 0.03,
    p = 3,
    d_max = 0.10
  ),
  scenario = "baseline",
  seed = 42
)

daily_saba <- daily_list$Saba
head(daily_saba)
```

Summarize the daily output. This gives a quick check on how many tropical-storm and hurricane days appear in the selected simulated years.

```{r}
#| label: summarize-daily-output
daily_saba |>
  summarise(
    total_days = n(),
    tc_days = sum(wind_kt >= 34, na.rm = TRUE),
    hur_days = sum(wind_kt >= 64, na.rm = TRUE),
    mean_annual_damage = sum(damage_rate, na.rm = TRUE) / n_distinct(sim_year)
  )
```

## Practical Notes {#sec-notes}

Keep these points in mind when you set up your own runs:

| Topic | Guidance |
|---|---|
| Baseline climate | Use `make_climate_cfg(scenario = "stationary")`. There is no longer a separate climate-off mode. |
| Hypothetical warming | Use `make_climate_cfg(delta_sst = ...)` when you want a transparent stress-test input in degrees C. |
| Scenario time slice | Use `make_climate_cfg(scenario = "ssp585", target_year = ...)` when you want named-scenario provenance and automatic `delta_sst` resolution. |
| Hazard config | Put climate inside `make_hazard_cfg(...)`; `run_hazard_model()` reads it from `cfg`. |
| Daily impacts | Pass a `damage` list, not legacy damage arguments, and extract the requested location from the returned named list. |

## Summary {#sec-summary}

You now have the current minimal workflow for the hazard model. The key pattern is that climate is always carried inside `hazard_cfg`, whether you are running a stationary baseline, a direct `delta_sst` stress test, or a scenario-based `target_year` run.

From here, the usual next step is to run more targets, increase `simulation_years`, or pass the daily hazard output into downstream impact or system analyses.
