---
title: "Tutorial 2 — Multi-Site Baseline & Climate Change Analysis"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

::: callout-note
## Prerequisites
This tutorial assumes you have already worked through [Tutorial 1](tutorial_1_introduction_setup.qmd), where the single-site workflow and the main output objects are introduced in more detail.
:::

This tutorial shows how to move from one island to a small group of islands, then compare a stationary baseline with a warmer-climate scenario.
The focus is practical: run the model, inspect the annual outputs, and then downscale one site to daily disruption metrics.

## What This Tutorial Covers {#sec-what-this-covers}

By the end, you will be able to:

- run the baseline model for three islands at once
- compare a stationary baseline with a warmer climate scenario
- translate annual scenario differences into daily disruption metrics for one site

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
library(ggplot2)
```

## Multi-Site Baseline {#sec-multisite-baseline}

We start with the three Dutch Caribbean islands used throughout this package: St. Martin, Saba, and Statia.
The setup is the same as in Tutorial 1, except that `targets` now contains three rows instead of one.

```{r}
#| label: setup-baseline
seed <- 123L
simulation_years <- 500L

ibtracs_demo <- system.file(
  "extdata",
  "ibtracs_demo.csv",
  package = "ipdcstorm"
)

targets <- data.frame(
  name = c("St_Martin", "Saba", "Statia"),
  lat = c(18.0708, 17.6350, 17.4890),
  lon = c(-63.0501, -63.2300, -62.9740)
)

cfg_baseline <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  search_radius_km = 800,
  simulation_years = simulation_years,
  climate = make_climate_cfg(scenario = "stationary")
)
```

```{r}
#| label: run-baseline
out_baseline <- run_hazard_model(
  cfg = cfg_baseline,
  targets = targets,
  seed = seed,
  verbose = FALSE
)
```

The `rates` table is the historical calibration result: mean annual site-level rates for tropical storms and hurricanes at each island.

```{r}
#| label: inspect-baseline-rates
out_baseline$rates
```

The annual simulation table gives one synthetic season per row.
For a quick comparison across islands, we can average the simulated counts and hurricane share by location.

```{r}
#| label: summarise-baseline-sim
baseline_summary <- out_baseline$sim |>
  group_by(location) |>
  summarise(
    mean_total = mean(n_total, na.rm = TRUE),
    mean_ts = mean(n_ts, na.rm = TRUE),
    mean_hur = mean(n_hur, na.rm = TRUE),
    mean_p_hur = mean(p_hurricane, na.rm = TRUE),
    .groups = "drop"
  )

baseline_summary
```

| Column | Description |
|--------|-------------|
| `mean_total` | Mean annual number of site-impacting storms with at least tropical-storm strength. |
| `mean_ts` | Mean annual number of tropical-storm events. |
| `mean_hur` | Mean annual number of hurricane events. |
| `mean_p_hur` | Mean simulated fraction of storms that are hurricanes. |

These averages should broadly line up with the rate table.
Small differences across islands reflect real differences in how often storms pass close enough, and strongly enough, to exceed the site thresholds.

## Climate Change Scenarios {#sec-climate-change}

The climate extension changes the annual simulation in two SST-driven ways.
First, warming can scale the total number of storms up or down through a rate effect.
Second, warming can shift the storm mix toward a larger hurricane fraction through an intensity effect.

In this model, both effects are driven by a scalar `delta_sst`: the Main Development Region sea-surface temperature change, in °C, relative to a 1991-2020 baseline.
That keeps the workflow simple for scenario testing: you can either use a named scenario helper or supply a direct warming value yourself.

We will compare a stationary baseline (`delta_sst = 0`) with a hypothetical warmer climate using `delta_sst = 1.5`.
That is a clean side-by-side example for understanding the direction and size of the climate response.

```{r}
#| label: inspect-sst-scenarios
sst_scenario_info() |>
  filter(source == "ipcc_ar6") |>
  select(scenario, source, delta_sst_2050, delta_sst_2100)
```

The built-in scenario table is useful for context, but for this tutorial we set the warming directly so the climate signal is explicit in the code.

```{r}
#| label: configure-future-climate
climate_future <- make_climate_cfg(
  delta_sst = 1.5,
  sensitivity_mode = "fixed"
)

cfg_future <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  search_radius_km = 800,
  simulation_years = simulation_years,
  climate = climate_future
)
```

```{r}
#| label: run-future-climate
out_future <- run_hazard_model(
  cfg = cfg_future,
  targets = targets,
  seed = seed,
  verbose = FALSE
)
```

Now we stack the two annual simulation tables and compute a compact comparison by scenario and island.
The summary below reports the mean annual storm count, the tropical-storm and hurricane components, and the mean simulated hurricane fraction.

```{r}
#| label: summarise-scenarios
activity_summary <- bind_rows(
  out_baseline$sim |> mutate(scenario = "Baseline"),
  out_future$sim |> mutate(scenario = "Future (+1.5 C)")
) |>
  group_by(scenario, location) |>
  summarise(
    mean_total = mean(n_total, na.rm = TRUE),
    mean_ts = mean(n_ts, na.rm = TRUE),
    mean_hur = mean(n_hur, na.rm = TRUE),
    mean_p_hur = mean(p_hurricane, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(location, scenario)

activity_summary
```

| Column | Description |
|--------|-------------|
| `scenario` | Baseline or future climate label used for the comparison. |
| `location` | Target island. |
| `mean_total` | Mean annual storm count under that scenario. |
| `mean_ts` | Mean annual tropical-storm count. |
| `mean_hur` | Mean annual hurricane count. |
| `mean_p_hur` | Mean simulated hurricane fraction. |

```{r}
#| label: plot-scenario-activity
activity_summary |>
  ggplot(aes(x = location, y = mean_total, fill = scenario)) +
  geom_col(position = "dodge") +
  labs(
    x = NULL,
    y = "Mean annual storm count",
    fill = "Scenario",
    title = "Annual storm activity by scenario and location"
  ) +
  theme_minimal(base_size = 11)
```

::: callout-tip
## Interpreting scenario differences
These scenario differences are changes in long-run stochastic behavior, not forecasts for any single year. A higher `mean_total` or `mean_p_hur` means the simulated climate state is more active on average, but quiet years still occur.
:::

## Daily Series Comparison {#sec-daily-comparison}

Annual counts are useful for climatology, but many planning questions depend on daily disruption.
To illustrate that step, we generate daily hazard-impact series for Saba under the baseline and the warmer scenario.

```{r}
#| label: generate-daily-series
daily_baseline <- generate_daily_hazard_impact_spatial(
  out = out_baseline,
  location = "Saba",
  sim_years = 1:200,
  year0 = 2025,
  gust_factor = 1.25,
  damage = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario = "Baseline",
  seed = 123
)$Saba

daily_future <- generate_daily_hazard_impact_spatial(
  out = out_future,
  location = "Saba",
  sim_years = 1:200,
  year0 = 2025,
  gust_factor = 1.25,
  damage = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario = "Future (+1.5 C)",
  seed = 123
)$Saba
```

The combined table lets us compare how often Saba experiences tropical-storm and hurricane days, along with the average annual damage-rate signal over the 200 simulated years.

```{r}
#| label: summarise-daily-series
daily_summary <- bind_rows(daily_baseline, daily_future) |>
  group_by(scenario) |>
  summarise(
    tc_days = sum(wind_kt > 0, na.rm = TRUE),
    ts_days = sum(wind_kt >= 34, na.rm = TRUE),
    hur_days = sum(wind_kt >= 64, na.rm = TRUE),
    mean_annual_damage = sum(damage_rate, na.rm = TRUE) / 200,
    .groups = "drop"
  )

daily_summary
```

```{r}
#| label: plot-daily-comparison
daily_plot_data <- data.frame(
  scenario = rep(daily_summary$scenario, times = 3),
  metric = rep(
    c("TC days/year", "TS days/year", "HUR days/year"),
    each = nrow(daily_summary)
  ),
  value = c(
    daily_summary$tc_days / 200,
    daily_summary$ts_days / 200,
    daily_summary$hur_days / 200
  )
)

daily_plot_data |>
  ggplot(aes(x = metric, y = value, fill = scenario)) +
  geom_col(position = "dodge") +
  labs(
    x = NULL,
    y = "Mean annual days",
    fill = "Scenario",
    title = "Daily disruption metrics for Saba by scenario"
  ) +
  theme_minimal(base_size = 11)
```

| Column | Description |
|--------|-------------|
| `tc_days` | Number of simulated days with any tropical-cyclone wind signal (`wind_kt > 0`). |
| `ts_days` | Number of days reaching at least 34 kt. |
| `hur_days` | Number of days reaching at least 64 kt. |
| `mean_annual_damage` | Average annual damage-rate total across the 200 simulated years. |

::: callout-note
## About Level 3 perturbation
The package also supports Level 3 storm perturbation for more exploratory scenario analysis. We do not use it here, so the daily comparison isolates the Level 1 and Level 2 climate effects discussed above.
:::

## Key Parameters {#sec-key-parameters}

For most climate sensitivity runs, three settings matter most:

| Parameter | What it controls |
|-----------|------------------|
| `delta_sst` | The imposed MDR SST shift in °C relative to the 1991-2020 baseline. In this tutorial, `delta_sst = 1.5` defines the future climate state directly. |
| `scenario` vs `delta_sst` | Use a named `scenario` when you want a built-in helper tied to a source such as IPCC AR6. Use `delta_sst` when you want a transparent what-if value without committing to one named pathway. |
| `sensitivity_mode` | `"fixed"` keeps the calibrated climate sensitivities unchanged as warming increases. `"linear_shifted"` lets those sensitivities themselves vary with `delta_sst`, which is more exploratory. |

This baseline-versus-scenario pattern is a good starting point for stress testing.
Once you are comfortable with the annual and daily outputs, you can expand the same workflow to additional islands or alternative warming assumptions.
