---
title: "Tutorial 2 — Model Validation"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

::: callout-note
## Prerequisites
This tutorial assumes you have completed [Tutorial 1](tutorial_1_setup.qmd), which introduces the baseline hazard workflow and the `out` object that we validate here.
:::

This tutorial mirrors the production validation pipeline in `inst/extcode/02-model-validation.R`.
We run a stationary baseline simulation, exercise the three-tier validation suite, and inspect the diagnostics that result.

## What This Tutorial Covers {#sec-what-this-covers}

By the end, you will be able to:

- create a baseline model output suitable for validation;
- configure and run the validation suite with `make_validation_cfg()` and `run_validation_suite()`;
- interpret each of the three validation tiers (hindcast return levels, rate sanity, wind-field spot-checks);
- read the saved plots and tables produced when `save_plots = TRUE` and `save_tables = TRUE`.

## Overview {#sec-overview}

The validation framework has three tiers, run in one suite call:

1. **Tier 1 — Hindcast.** Hold out the last `HOLDOUT_YEARS` of the historical record, refit on the training period, simulate `N_SIM` synthetic years, and compare simulated vs. observed return levels at the requested return periods. A well-calibrated model places simulated return levels inside the observed confidence interval at most sites.
2. **Tier 2 — Rate check.** Compare calibrated annual storm-class rates (TS, HUR) against published HURDAT2/IBTrACS climatologies, after adjusting for the methodological mismatch between site-exceedance counting and basin-track counting.
3. **Tier 3 — Wind field.** Spot-check the Holland wind-profile model against historical station observations for documented storm passages and report site-level bias.

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
library(ggplot2)
```

## Configuration {#sec-config}

Settings here match the production script `inst/extcode/02-model-validation.R`.
Lower `N_SIM` to render the tutorial faster while you are exploring; use 2,000 for any defensible result.

```{r}
#| label: config
SEED             <- 42L
N_SIM            <- 2000L
HIST_START_YEAR  <- 1970L
SEARCH_RADIUS_KM <- 800

# Years held out from the end of the historical record for the hindcast.
HOLDOUT_YEARS <- 10L

# Return periods (yr) to evaluate. Keep these within ~ record length / 3
# so that empirical observed quantiles remain stable.
RETURN_PERIODS <- c(5, 10, 25, 50, 100)

# Confidence level used for the observed return-level CI in Tier 1.
CONF_LEVEL <- 0.95

# Where saved plots/tables go.
SAVE_OUTPUTS <- TRUE
OUT_DIR      <- "output/validation"
```

```{r}
#| label: targets
targets <- tibble::tribble(
  ~name,         ~lat,      ~lon,
  "St_Martin",   18.0708,  -63.0501,
  "Saba",        17.6350,  -63.2300,
  "Statia",      17.4890,  -62.9740
)
```

## Create a Model Output to Validate {#sec-create-output}

The validation suite operates on a standard `run_hazard_model()` output.
We use the packaged 1970+ IBTrACS subset so the tutorial renders end-to-end without external data.

```{r}
#| label: hazard-config
DATA_PATH <- system.file("extdata", "ibtracs_1970.csv", package = "ipdcstorm")

hazard_cfg <- make_hazard_cfg(
  data_path             = DATA_PATH,
  search_radius_km      = SEARCH_RADIUS_KM,
  historical_start_year = HIST_START_YEAR,
  simulation_years      = N_SIM,
  climate               = make_climate_cfg(scenario = "stationary")
)
```

```{r}
#| label: hazard-run
out_base <- run_hazard_model(
  cfg     = hazard_cfg,
  targets = targets,
  seed    = SEED,
  verbose = FALSE
)
```

## Validation Configuration {#sec-validation-config}

`make_validation_cfg()` packages the validation parameters and the output behaviour into one object.
Setting `full_output = TRUE` returns the internal hindcast object and the file paths of saved artifacts in addition to the slim summary tibbles.

```{r}
#| label: validation-config
val_cfg <- make_validation_cfg(
  holdout_years  = HOLDOUT_YEARS,
  n_sim          = N_SIM,
  return_periods = RETURN_PERIODS,
  conf_level     = CONF_LEVEL,
  seed           = SEED,
  out_dir        = OUT_DIR,
  save_plots     = SAVE_OUTPUTS,
  save_tables    = SAVE_OUTPUTS,
  full_output    = TRUE
)

val_cfg
```

| Parameter | Description |
|-----------|-------------|
| `holdout_years` | Number of years withheld from the end of the historical record for the train/test split. |
| `n_sim` | Number of synthetic years simulated for the hindcast. `NULL` inherits from `out_base`. |
| `return_periods` | Return periods (yr) at which observed and simulated return levels are compared. |
| `conf_level` | Confidence level for the observed return-level CI; controls the strictness of the Tier 1 pass test. |
| `seed` | RNG seed; should match the seed used for the baseline run. |
| `out_dir` | Destination directory for saved plots and tables. |
| `save_plots` / `save_tables` | Toggles for writing PNG figures and CSV tables. |
| `full_output` | When `TRUE`, also returns `hindcast`, `climate_response`, and `artifacts` (file paths). |

## Run the Validation Suite {#sec-run-validation}

A single call runs all three tiers and returns a structured list.
The function prints a console summary as it works.

```{r}
#| label: run-validation
val_out <- run_validation_suite(
  out = out_base,
  cfg = val_cfg
)
```

```{r}
#| label: validation-summary
val_out$summary
```

::: callout-note
## What's in `val_out`
Default output contains `summary`, `comparison` (Tier 1), `rate_check` (Tier 2), `wind_field` (Tier 3), and `plots` (list of ggplot objects). With `full_output = TRUE`, you also get `hindcast`, `climate_response`, and `artifacts` (paths to saved PNGs and CSVs). The on-disk bundle lives in `OUT_DIR`.
:::

## Tier 1 — Hindcast Return Levels {#sec-tier1}

Tier 1 asks whether the model, refit on the training record, reproduces return levels implied by the full observed record.
The pass test is whether `sim_rl` falls inside the observed `[obs_ci_lo, obs_ci_hi]` interval.

```{r}
#| label: tier1-table
hc_tbl <- val_out$comparison

hc_tbl |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  arrange(location, return_period)
```

```{r}
#| label: tier1-passrate
n_checks  <- nrow(hc_tbl)
n_in_ci   <- sum(hc_tbl$obs_in_ci, na.rm = TRUE)
pct_in_ci <- round(100 * n_in_ci / n_checks, 1)

sprintf("Pass rate: %d / %d checks within %d%% CI  (%.1f%%)",
        n_in_ci, n_checks, round(CONF_LEVEL * 100), pct_in_ci)
```

| Column | Description |
|--------|-------------|
| `location` | Target site. |
| `return_period` | Return period in years. |
| `obs_full_rl` | Observed return level from the full historical record (kt). |
| `sim_rl` | Simulated return level from the `N_SIM` ensemble (kt). |
| `obs_ci_lo` / `obs_ci_hi` | Bounds of the observed return-level CI at `conf_level` (kt). |
| `obs_in_ci` | `TRUE` if `sim_rl` falls within the observed CI. |
| `bias_pct` | Percentage bias `100 * (sim_rl - obs_full_rl) / obs_full_rl`; positive = overprediction. |

```{r}
#| label: fig-hindcast
#| fig-cap: "Hindcast validation: simulated vs observed return levels."
val_out$plots$hindcast_return_levels
```

```{r}
#| label: fig-qq
#| fig-cap: "QQ plot of simulated vs observed annual maxima."
val_out$plots$qq_annual_max
```

```{r}
#| label: fig-cdf
#| fig-cap: "CDF comparison of simulated and observed annual maxima."
val_out$plots$exceedance_comparison
```

::: callout-note
## Why the CI is on the observed curve
CIs are placed on the observed return levels, not the simulated ones. Once the synthetic sample is large, fitting uncertainty on `sim_rl` is small relative to sampling uncertainty in the short observational record.
:::

::: callout-tip
## Read the QQ plot first
The QQ plot is the fastest way to localize bias by intensity. Upper-tail points above the 1:1 line indicate the model is overshooting the tail; points below indicate underprediction.
:::

## Tier 2 — Rate Sanity Check {#sec-tier2}

Tier 2 compares the model's annual rates of TS- and HUR-class impacts against published climatology.
The model counts site exceedances; references count basin-track passages within a radius.
The check applies an expected-ratio adjustment before flagging.

```{r}
#| label: tier2-table
val_out$rate_check |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

```{r}
#| label: tier2-flagged
n_flagged <- sum(val_out$rate_check$flag != "OK", na.rm = TRUE)

if (n_flagged > 0) {
  sprintf("%d rate(s) flagged — review search_radius_km or historical_start_year.", n_flagged)
} else {
  "All rates within reference bounds."
}
```

| Column | Description |
|--------|-------------|
| `lambda_model_raw` | Raw annual rate estimated by the model before adjustment. |
| `lambda_ref` | Published reference annual rate for the site/storm class. |
| `lambda_adj` | Adjusted model rate used for the comparison. |
| `expected_ratio` | Expected model/reference ratio given counting-method differences. |
| `adj_ratio` | Adjusted model-to-reference ratio (after `expected_ratio` correction). |
| `flag` | `OK`, `elevated`, `slightly_low`, `HIGH`, or `LOW`. |

::: callout-note
## Flag interpretation
`OK` is within band. `elevated` / `slightly_low` are mild discrepancies worth noting. `HIGH` / `LOW` are stronger mismatches: investigate `search_radius_km`, `historical_start_year`, or possible regional rate departures from basin climatology.
:::

## Tier 3 — Wind-Field Spot-Checks {#sec-tier3}

Tier 3 compares modelled site winds for individual historical storms against station observations.
Observations are graded by quality class with class-specific bias tolerances:
`A` = direct, ±15 %; `B` = converted, ±25 %; `C` = approximate, ±35 %.

```{r}
#| label: tier3-table
val_out$wind_field |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))
```

| Column | Description |
|--------|-------------|
| `storm_name` | Historical storm used for the comparison. |
| `location` | Target site. |
| `obs_1min_equiv_kt` | Observed wind, converted to a 1-minute sustained equivalent (kt). |
| `model_V_site_kt` | Modelled peak site wind (kt). |
| `bias_kt` | Model − observed (kt). |
| `bias_pct` | Percentage bias. |
| `obs_quality` | Observation quality class (`A`, `B`, `C`). |
| `bias_ok` | `TRUE` if `|bias_pct|` falls within the class tolerance. |

```{r}
#| label: fig-wind-field
#| fig-cap: "Wind-field validation: modelled vs observed site winds."
val_out$plots$wind_field_scatter
```

```{r}
#| label: fig-bias-decomposition
#| fig-cap: "Bias decomposition: frequency vs intensity contributions."
val_out$plots$bias_decomposition
```

::: callout-note
## This tier is the hardest to pass cleanly
Station observations are point measurements affected by terrain, exposure, and instrumentation. Biases of order ±20 % are expected; persistent same-sign bias across many storms suggests the radius–wind relationship needs regional tuning.
:::

## Interpreting Results and Next Steps {#sec-interpretation}

Read the tiers together rather than in isolation:

- A **Tier 1** miss (sim outside CI) without a Tier 2 or Tier 3 problem usually points to tail-shape issues — check the GEV fit, holdout length, and whether the most intense observed events fall in the holdout.
- A **Tier 2** flag normally points to a counting/scope mismatch — re-examine `search_radius_km`, `historical_start_year`, and the per-class reference ratios.
- A **Tier 3** persistent bias points to the wind-profile parameterisation (Holland B, RMW estimate, asymmetry) rather than to the stochastic generator.

Use the bias decomposition plot to separate frequency-driven bias (too many or too few impactful events) from intensity-driven bias (events of the right count but wrong magnitude).

::: callout-tip
## One-step shortcut
`validate_hazard_model()` runs the hazard model and the validation suite in one call. Keep the two-step pattern shown here while you are inspecting outputs; switch to the wrapper for repeat runs.
:::

::: callout-tip
## Targeted reruns
The package also exposes the individual tier functions: `validate_rates()` and `validate_wind_field()`. Use these to iterate on a single tier without re-running the others.
:::

## Saved Outputs {#sec-saved-outputs}

With `save_plots = TRUE` and `save_tables = TRUE`, the suite writes a small report bundle to `OUT_DIR`.

```{r}
#| label: saved-artifacts
val_out$artifacts
```

| File | Description |
|------|-------------|
| `hindcast_return_levels.png` | Return-level comparison with confidence intervals. |
| `qq_annual_max.png` | QQ plot of annual maxima. |
| `exceedance_comparison.png` | CDF comparison of observed and simulated annual maxima. |
| `wind_field_scatter.png` | Modelled vs observed site winds. |
| `bias_decomposition.png` | Frequency vs intensity bias contributions. |
| `hindcast_return_levels.csv` | Tier 1 comparison table. |
| `rate_check.csv` | Tier 2 rate-check table. |
| `wind_field.csv` | Tier 3 wind-field bias table. |
| `validation_summary.csv` | Compact per-location validation summary across tiers. |

Once you can read the three tiers fluently, the same workflow scales to your own sites and longer model runs.
A model that fails Tier 1 at more than one site should be recalibrated (revisit `search_radius_km` or `historical_start_year`) before drawing conclusions from any climate-change projection built on top of it.
