---
title: "Tutorial 3 - Climate Impact Analysis"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

## 1. What This Tutorial Covers {#sec-what-this-covers}

This tutorial scales the workflow from a single site to a small group of islands, then compares a present-day baseline with a warmer climate.

By the end, you will be able to:

- run the baseline model for three islands at once;
- compare today's climate with a warmer scenario;
- turn the annual scenario differences into day-by-day disruption metrics for one site.

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
library(ggplot2)
```

## 2. Multi-Site Baseline {#sec-multisite-baseline}

We start with the three Dutch Caribbean islands used throughout the package: St. Martin, Saba, and Statia.

```{r}
#| label: setup-baseline
seed <- 123L
simulation_years <- 500L

ibtracs_demo <- system.file(
  "extdata",
  "ibtracs_demo.csv",
  package = "ipdcstorm"
)

targets <- data.frame(
  name = c("St_Martin", "Saba", "Statia"),
  lat = c(18.0708, 17.6350, 17.4890),
  lon = c(-63.0501, -63.2300, -62.9740)
)

cfg_baseline <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  search_radius_km = 800,
  simulation_years = simulation_years,
  climate = make_climate_cfg(scenario = "stationary")
)
```

```{r}
#| label: run-baseline
out_baseline <- run_hazard_model(
  cfg = cfg_baseline,
  targets = targets,
  seed = seed,
  verbose = FALSE
)
```

The `rates` table shows what the model learned from history: the average number of tropical-storm and hurricane hits per year at each island.

```{r}
#| label: inspect-baseline-rates
out_baseline$rates
```

The simulation table has one synthetic season per row.
To compare islands at a glance, average the simulated counts and the hurricane share by location.

```{r}
#| label: summarise-baseline-sim
baseline_summary <- out_baseline$sim |>
  group_by(location) |>
  summarise(
    mean_total = mean(n_total, na.rm = TRUE),
    mean_ts = mean(n_ts, na.rm = TRUE),
    mean_hur = mean(n_hur, na.rm = TRUE),
    mean_p_hur = mean(p_hurricane, na.rm = TRUE),
    .groups = "drop"
  )

baseline_summary
```

| Column | Description |
|--------|-------------|
| `mean_total` | Average number of storms per year that hit the site at tropical-storm strength or above. |
| `mean_ts` | Average number of tropical-storm events per year. |
| `mean_hur` | Average number of hurricanes per year. |
| `mean_p_hur` | Average share of events that reach hurricane strength. |

## 3. Climate Change Scenarios {#sec-climate-change}

Warming affects the simulated climate in two ways.
First, it can raise or lower the total number of storms per year (a *rate* effect).
Second, it can shift the storm mix so a larger share reach hurricane strength (an *intensity* effect).

Both effects are driven by a single number: `delta_sst`, the change in Main Development Region sea-surface temperature (in °C) relative to a 1991-2020 baseline.
You can either use a named scenario helper (IPCC AR6, KNMI'23) or set `delta_sst` directly.

Here we compare a stationary baseline (`delta_sst = 0`) with a warmer climate at `delta_sst = 1.5`.

```{r}
#| label: inspect-sst-scenarios
sst_scenario_info() |>
  filter(source == "ipcc_ar6") |>
  select(scenario, source, delta_sst_2050, delta_sst_2100)
```

The built-in scenario table is useful context, but for the tutorial we set `delta_sst` directly so the warming is visible in the code.

```{r}
#| label: configure-future-climate
climate_future <- make_climate_cfg(
  delta_sst = 1.5,
  sensitivity_mode = "fixed"
)

cfg_future <- make_hazard_cfg(
  data_path = ibtracs_demo,
  historical_start_year = 1970L,
  search_radius_km = 800,
  simulation_years = simulation_years,
  climate = climate_future
)
```

```{r}
#| label: run-future-climate
out_future <- run_hazard_model(
  cfg = cfg_future,
  targets = targets,
  seed = seed,
  verbose = FALSE
)
```

Now stack the two simulation tables and compare by scenario and island.
The summary reports the average yearly storm count, how it splits into tropical storms and hurricanes, and the simulated hurricane share.

```{r}
#| label: summarise-scenarios
activity_summary <- bind_rows(
  out_baseline$sim |> mutate(scenario = "Baseline"),
  out_future$sim |> mutate(scenario = "Future (+1.5 C)")
) |>
  group_by(scenario, location) |>
  summarise(
    mean_total = mean(n_total, na.rm = TRUE),
    mean_ts = mean(n_ts, na.rm = TRUE),
    mean_hur = mean(n_hur, na.rm = TRUE),
    mean_p_hur = mean(p_hurricane, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(location, scenario)

activity_summary
```

| Column | Description |
|--------|-------------|
| `scenario` | Baseline or future — the label used in the comparison. |
| `location` | Target island. |
| `mean_total` | Average yearly storm count under that scenario. |
| `mean_ts` | Average yearly tropical-storm count. |
| `mean_hur` | Average yearly hurricane count. |
| `mean_p_hur` | Average share of storms reaching hurricane strength. |

```{r}
#| label: plot-scenario-activity
activity_summary |>
  ggplot(aes(x = location, y = mean_total, fill = scenario)) +
  geom_col(position = "dodge") +
  labs(
    x = NULL,
    y = "Mean annual storm count",
    fill = "Scenario",
    title = "Annual storm activity by scenario and location"
  ) +
  theme_minimal(base_size = 11)
```

## 4. Daily Series Comparison {#sec-daily-comparison}

Annual counts are useful for climatology, but many planning questions are about daily disruption.
To show that step, we generate daily hazard and impact series for Saba under the baseline and the warmer scenario.

```{r}
#| label: generate-daily-series
daily_baseline <- generate_daily_hazard_impact_spatial(
  out = out_baseline,
  location = "Saba",
  sim_years = 1:200,
  year0 = 2025,
  gust_factor = 1.25,
  damage = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario = "Baseline",
  seed = 123
)$Saba

daily_future <- generate_daily_hazard_impact_spatial(
  out = out_future,
  location = "Saba",
  sim_years = 1:200,
  year0 = 2025,
  gust_factor = 1.25,
  damage = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario = "Future (+1.5 C)",
  seed = 123
)$Saba
```

The combined table shows how often Saba experiences tropical-storm and hurricane days under each scenario, plus the average annual damage signal across the 200 simulated years.

```{r}
#| label: summarise-daily-series
daily_summary <- bind_rows(
  dplyr::mutate(daily_baseline, scenario = "Baseline"),
  dplyr::mutate(daily_future, scenario = "Future (+1.5 C)")
) |>
  group_by(scenario) |>
  summarise(
    tc_days = sum(wind_kt > 0, na.rm = TRUE),
    ts_days = sum(wind_kt >= 34, na.rm = TRUE),
    hur_days = sum(wind_kt >= 64, na.rm = TRUE),
    mean_annual_damage = sum(damage_rate, na.rm = TRUE) / 200,
    .groups = "drop"
  )

daily_summary
```

```{r}
#| label: plot-daily-comparison
daily_plot_data <- data.frame(
  scenario = rep(daily_summary$scenario, times = 3),
  metric = rep(
    c("TC days/year", "TS days/year", "HUR days/year"),
    each = nrow(daily_summary)
  ),
  value = c(
    daily_summary$tc_days / 200,
    daily_summary$ts_days / 200,
    daily_summary$hur_days / 200
  )
)

daily_plot_data |>
  ggplot(aes(x = metric, y = value, fill = scenario)) +
  geom_col(position = "dodge") +
  labs(
    x = NULL,
    y = "Mean annual days",
    fill = "Scenario",
    title = "Daily disruption metrics for Saba by scenario"
  ) +
  theme_minimal(base_size = 11)
```

| Column | Description |
|--------|-------------|
| `tc_days` | Days where a storm produced any wind at the site (`wind_kt > 0`). |
| `ts_days` | Days reaching at least 34 kt (tropical-storm strength). |
| `hur_days` | Days reaching at least 64 kt (hurricane strength). |
| `mean_annual_damage` | Average damage total per year across the 200 simulated seasons. |

## 5. Key Parameters {#sec-key-parameters}

For most climate sensitivity runs, three settings matter most:

| Parameter | What it controls |
|-----------|------------------|
| `delta_sst` | The warming (°C) relative to the 1991-2020 baseline. Here `delta_sst = 1.5` sets the future climate directly. |
| `scenario` vs `delta_sst` | Use a named `scenario` (e.g. IPCC AR6, KNMI'23) when you want to follow a published pathway. Use `delta_sst` when you want a transparent what-if value without committing to any specific pathway. |
| `sensitivity_mode` | `"fixed"` keeps the calibrated climate sensitivities constant as warming grows. `"linear_shifted"` lets those sensitivities themselves vary with `delta_sst` — more exploratory. |
