---
title: "Tutorial 2 - Model Validation"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

::: callout-note
## Prerequisites
This tutorial assumes you have completed [Tutorial 1](tutorial_1_setup.qmd), which introduces the baseline hazard workflow and the `out` object that we validate here.
:::

## 1. Overview {#sec-overview}

Validation runs three checks in a single suite:

1. **Tier 1 — Hindcast.** Hide the most recent `HOLDOUT_YEARS` of the record, refit on the older part, simulate `N_SIM` synthetic years, and compare the model's return levels against what the full record.
2. **Tier 2 — Rate check.** Compare the model's annual storm rates (TS, HUR) against published climatology. 
3. **Tier 3 — Wind field.** For a handful of historical storms with station records, check whether the modelled wind at each site matches what was actually observed.

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
library(ggplot2)
```

## 2. Configuration {#sec-config}

These settings match the production script.
Lower `N_SIM` while you are exploring; use 2,000 or more for any result you plan to act on.

```{r}
#| label: config
SEED             <- 42L
N_SIM            <- 2000L
HIST_START_YEAR  <- 1970L
SEARCH_RADIUS_KM <- 800

# Years held out from the end of the historical record for the hindcast.
HOLDOUT_YEARS <- 10L

# Return periods (yr) to evaluate
RETURN_PERIODS <- c(5, 10, 25, 50, 100)

# Confidence level used for the observed return-level CI in Tier 1.
CONF_LEVEL <- 0.95

# Where saved plots/tables go.
SAVE_OUTPUTS <- TRUE
OUT_DIR      <- "output/validation"
```

```{r}
#| label: targets
targets <- tibble::tribble(
  ~name,         ~lat,      ~lon,
  "St_Martin",   18.0708,  -63.0501,
  "Saba",        17.6350,  -63.2300,
  "Statia",      17.4890,  -62.9740
)
```

## 3. Create a Model Configuration {#sec-create-output}

`validate_hazard_model()` runs the hazard model and validation together.
We still spell out the hazard configuration here so the model inputs are visible.
The packaged 1970+ IBTrACS subset keeps the example self-contained.

```{r}
#| label: hazard-config
DATA_PATH <- system.file("extdata", "ibtracs_1970.csv", package = "ipdcstorm")

hazard_cfg <- make_hazard_cfg(
  data_path             = DATA_PATH,
  search_radius_km      = SEARCH_RADIUS_KM,
  historical_start_year = HIST_START_YEAR,
  simulation_years      = N_SIM,
  climate               = make_climate_cfg(scenario = "stationary")
)
```

## 4. Validation Configuration {#sec-validation-config}

`make_validation_cfg()` bundles every validation option — how long the holdout is, which return periods to evaluate, where to write output — into one object.
Setting `full_output = TRUE` also hands back the internal hindcast object and the file paths of every saved artifact, on top of the compact summary tables.

```{r}
#| label: validation-config
val_cfg <- make_validation_cfg(
  holdout_years  = HOLDOUT_YEARS,
  n_sim          = N_SIM,
  return_periods = RETURN_PERIODS,
  conf_level     = CONF_LEVEL,
  seed           = SEED,
  out_dir        = OUT_DIR,
  save_plots     = SAVE_OUTPUTS,
  save_tables    = SAVE_OUTPUTS,
  full_output    = TRUE
)

val_cfg
```

| Parameter | Description |
|-----------|-------------|
| `holdout_years` | How many recent years to hide from the model and then use to test it. |
| `n_sim` | How many synthetic years to run for the hindcast. `NULL` reuses the number from `out_base`. |
| `return_periods` | Return periods (years) where observed and simulated values are compared. |
| `conf_level` | Width of the observed return-level confidence interval; stricter = harder to pass Tier 1. |
| `seed` | Random seed; use the same value as the baseline run for a fair comparison. |
| `out_dir` | Where to write saved plots and tables. |
| `save_plots` / `save_tables` | On/off switches for writing PNG figures and CSV tables. |
| `full_output` | When `TRUE`, also returns the richer `hindcast`, `climate_response`, and `artifacts` data. |

## 5. Run the Validation Workflow {#sec-run-validation}

One call runs the hazard model, then the three validation tiers, and returns both.
The function prints a short console summary as it goes.

```{r}
#| label: run-validation
validation_result <- validate_hazard_model(
  cfg = hazard_cfg,
  targets = targets,
  validation_cfg = val_cfg
)

out_base <- validation_result$out
val_out <- validation_result$val
```

```{r}
#| label: validation-summary
val_out$summary
```

::: callout-note
## What's in `validation_result`
The wrapper returns `out` (the hazard model output), `val` (the validation results), and `artifacts` (file paths if you turned on saving). Inside `val` you will find `summary`, `comparison` (Tier 1), `rate_check` (Tier 2), `wind_field` (Tier 3), and `plots`. With `full_output = TRUE`, `val` also carries more detailed hindcast diagnostics.
:::

## 6. Tier 1 — Hindcast Return Levels {#sec-tier1}

Tier 1 asks a simple question: if we retrain the model on an earlier slice of the record, does it reproduce the return levels seen in the full record?
The pass test is whether the simulated return level (`sim_rl`) lands inside the observed confidence interval `[obs_ci_lo, obs_ci_hi]`.

```{r}
#| label: tier1-table
hc_tbl <- val_out$comparison

hc_tbl |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  arrange(location, return_period)
```

```{r}
#| label: tier1-passrate
n_checks  <- nrow(hc_tbl)
n_in_ci   <- sum(hc_tbl$obs_in_ci, na.rm = TRUE)
pct_in_ci <- round(100 * n_in_ci / n_checks, 1)

sprintf("Pass rate: %d / %d checks within %d%% CI  (%.1f%%)",
        n_in_ci, n_checks, round(CONF_LEVEL * 100), pct_in_ci)
```

| Column | Description |
|--------|-------------|
| `location` | Target site. |
| `return_period` | Return period in years. |
| `obs_full_rl` | Observed return level from the full record (kt). |
| `sim_rl` | Simulated return level from the `N_SIM` ensemble (kt). |
| `obs_ci_lo` / `obs_ci_hi` | Lower and upper confidence bounds on the observed value (kt). |
| `obs_in_ci` | `TRUE` if the simulated value lands inside the observed CI. |
| `bias_pct` | Percentage bias. Positive means the model is overshooting. |

```{r}
#| label: fig-hindcast
#| fig-cap: "Hindcast validation: simulated vs observed return levels."
val_out$plots$hindcast_return_levels
```

```{r}
#| label: fig-qq
#| fig-cap: "QQ plot of simulated vs observed annual maxima."
val_out$plots$qq_annual_max
```

::: callout-note
## Why the CI is drawn on the observed curve
The uncertainty band is on the observed return levels, not the simulated ones. With a long synthetic record, the model's return levels are already well-pinned down — the short observational record is the bigger source of uncertainty.
:::

::: callout-tip
## Read the QQ plot first
The QQ plot is the fastest way to see where bias lives. Points above the diagonal mean the model overshoots the tail; points below mean it underpredicts.
:::

## 7. Tier 2 — Rate Sanity Check {#sec-tier2}

Tier 2 compares the model's average yearly count of tropical-storm and hurricane hits against published climatology.
The model counts times a site is exceeded; climatology counts how often a storm passes through a circle around the site.
The check adjusts for this counting difference before flagging anything.

```{r}
#| label: tier2-table
val_out$rate_check |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

```{r}
#| label: tier2-flagged
n_flagged <- sum(val_out$rate_check$flag != "OK", na.rm = TRUE)

if (n_flagged > 0) {
  sprintf("%d rate(s) flagged — review search_radius_km or historical_start_year.", n_flagged)
} else {
  "All rates within reference bounds."
}
```

| Column | Description |
|--------|-------------|
| `lambda_model_raw` | The model's raw yearly rate before adjustment. |
| `lambda_ref` | Published reference rate for the site and storm class. |
| `lambda_adj` | The model rate after the counting-method adjustment. |
| `expected_ratio` | How the model rate *should* compare to the reference given the method difference. |
| `adj_ratio` | The adjusted model-to-reference ratio — what actually gets checked. |
| `flag` | `OK`, `elevated`, `slightly_low`, `HIGH`, or `LOW`. |

::: callout-note
## What the flags mean
`OK` is inside the expected band. `elevated` / `slightly_low` are mild mismatches worth noting. `HIGH` / `LOW` are the serious ones — revisit `search_radius_km`, `historical_start_year`, or consider whether the site's rate genuinely differs from basin averages.
:::

## 8. Tier 3 — Wind-Field Spot-Checks {#sec-tier3}

Tier 3 compares the model's site wind for a handful of historical storms against what real weather stations actually recorded.
Observations are graded by how reliable they are and each grade gets a different tolerance:
`A` (direct measurement) = ±15 %, `B` (converted) = ±25 %, `C` (estimated) = ±35 %.

```{r}
#| label: tier3-table
val_out$wind_field |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))
```

| Column | Description |
|--------|-------------|
| `storm_name` | Historical storm being spot-checked. |
| `location` | Target site. |
| `obs_1min_equiv_kt` | What the station recorded, converted to a 1-minute sustained wind (kt). |
| `model_V_site_kt` | What the model said the site wind was (kt). |
| `bias_kt` | Model minus observed (kt). |
| `bias_pct` | Percentage bias. |
| `obs_quality` | Observation grade (`A`, `B`, `C`). |
| `bias_ok` | `TRUE` if `|bias_pct|` is within this grade's tolerance. |

```{r}
#| label: fig-wind-field
#| fig-cap: "Wind-field validation: modelled vs observed site winds."
val_out$plots$wind_field_scatter
```

```{r}
#| label: fig-bias-decomposition
#| fig-cap: "Bias decomposition: frequency vs intensity contributions."
val_out$plots$bias_decomposition
```

::: callout-note
## This tier is the hardest to pass cleanly
Station records are point measurements, shaped by local terrain, instrument setup, and exposure. Biases of ±20 % or so are normal. What matters is whether the bias is persistently one-sided across many storms — that suggests the wind-field settings need regional tuning.
:::

## 9. Saved Outputs {#sec-saved-outputs}

With `save_plots = TRUE` and `save_tables = TRUE`, the suite writes a small report bundle to `OUT_DIR`.

```{r}
#| label: saved-artifacts
val_out$artifacts
```

| File | Description |
|------|-------------|
| `hindcast_return_levels.png` | Return-level comparison with confidence intervals. |
| `qq_annual_max.png` | QQ plot of annual maxima. |
| `exceedance_comparison.png` | CDF comparison of observed and simulated annual maxima. |
| `wind_field_scatter.png` | Modelled vs observed site winds. |
| `bias_decomposition.png` | Frequency vs intensity bias contributions. |
| `hindcast_return_levels.csv` | Tier 1 comparison table. |
| `rate_check.csv` | Tier 2 rate-check table. |
| `wind_field.csv` | Tier 3 wind-field bias table. |
| `validation_summary.csv` | Compact per-location validation summary across tiers. |

