---
title: "Tutorial 1 — Introduction, Setup & Baseline Workflow"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

We introduce the model, set up the configuration, fit the stationary baseline, and generate the daily wind series for three Dutch Caribbean targets.

## Model Overview {#sec-overview}

`ipdcstorm` is a parametric tropical-cyclone wind hazard model.
It combines NOAA's IBTrACS historical track database with stochastic simulation to generate long synthetic records of daily site-level wind exposure.
The model was developed for climate stress testing of supply-chain infrastructure on the Dutch Caribbean islands (St. Martin, Saba, Statia) but works for any point location in the North Atlantic basin.

The workflow has three stages:

1. **Historical calibration.** Read IBTrACS, filter to track points within `search_radius_km` of each site, and compute site-level winds with a Holland (1980) parametric profile that uses directional R34/R50/R64 wind radii (with Knaff-style climatological fallbacks) and a forward-motion asymmetry term. Site events are aggregated and classified by peak site wind into TS (≥ 34 kt) and HUR (≥ 64 kt). Annual class rates and a Gamma-Poisson overdispersion parameter `k_hat` are estimated from the historical record.
2. **Stochastic simulation.** Draw synthetic seasons from a two-level process: a Gamma-distributed annual activity multiplier `A` reproduces the observed burstiness of seasons, then total counts and the TS / HUR split follow Poisson and binomial draws conditional on `A`. Climate forcing — when enabled in Tutorial 3 — enters via SST-driven changes to annual rates and the hurricane fraction, with optional event-level perturbation of intensity, size, and translation speed.
3. **Temporal downscaling.** For each simulated year, sample events from a severity-stratified empirical event library, place them on the calendar with empirical day-of-year distributions, spread each event over its duration with a smooth pulse, and combine overlapping events into a daily wind (kt) series. A damage proxy and supporting daily fields (gust, surge, dominant-event metadata) are computed from the daily wind.

::: callout-note
## Where this tutorial sits
This tutorial covers the stationary baseline. [Tutorial 2](tutorial_2_validation.qmd) checks model skill via the three-tier validation suite. [Tutorial 3](tutorial_3_climate_change.qmd) layers KNMI'23 and IPCC AR6 climate-change scenarios on top of the same workflow.
:::

## Prerequisites {#sec-prerequisites}

### Install

```{r}
#| label: install-packages
# install.packages("devtools")
# devtools::install_github("tanerumit/ipdcstorm")
```

### IBTrACS data {#sec-ibtracs}

Calibration uses NOAA's **International Best Track Archive for Climate Stewardship (IBTrACS)**, the most comprehensive global dataset of tropical cyclone tracks.
The model uses the **North Atlantic subset** (`ibtracs.NA.list.v04r01.csv`), with 6-hourly track points for every recorded Atlantic storm.

| IBTrACS field(s) | Used for |
|---|---|
| `SID`, `SEASON`, `BASIN` | Storm identification, year, basin filter |
| `USA_LAT`, `USA_LON`, `ISO_TIME` | 6-hourly track position |
| `USA_WIND` | Maximum sustained wind (kt) |
| `USA_R34_NE/SE/SW/NW`, `USA_R50_*`, `USA_R64_*` | Directional wind radii (nm) at 34, 50, 64 kt thresholds |
| `USA_PRES`, `USA_POCI` | Central pressure and outermost closed-isobar pressure |
| `USA_RMW`, `USA_ROCI` | Radius of maximum wind and outermost closed isobar |

Wind radii drive the spatial extent of the wind field. When they are missing (mostly pre-2004), the model substitutes Knaff et al. (2015) climatological estimates.

**Download:** [IBTrACS NCEI](https://www.ncei.noaa.gov/products/international-best-track-archive)

For this tutorial we use the packaged 1970+ subset so it renders end-to-end.

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
```

## Configuration {#sec-config}

Settings here match the production script `inst/extcode/01-baseline-analysis.R`.
Lower `N_SIM` while exploring; use ≥ 1,000 for any defensible result and 2,000+ for production.

```{r}
#| label: config
SEED             <- 42L
N_SIM            <- 2000L
HIST_START_YEAR  <- 1970L
SEARCH_RADIUS_KM <- 800

# Calendar year mapped to sim_year = 1 in the daily output.
YEAR0 <- 2000L

# 1.0 → 1-min sustained wind (use this for validation against IBTrACS).
# 1.3 → standard WMO gust factor (use for damage / infrastructure impact).
GUST_FACTOR <- 1.0
```

```{r}
#| label: targets
targets <- tibble::tribble(
  ~name,         ~lat,      ~lon,
  "St_Martin",   18.0708,  -63.0501,
  "Saba",        17.6350,  -63.2300,
  "Statia",      17.4890,  -62.9740
)
```

```{r}
#| label: hazard-cfg
DATA_PATH <- system.file("extdata", "ibtracs_1970.csv", package = "ipdcstorm")

hazard_cfg <- make_hazard_cfg(
  data_path             = DATA_PATH,
  search_radius_km      = SEARCH_RADIUS_KM,
  historical_start_year = HIST_START_YEAR,
  simulation_years      = N_SIM,
  climate               = make_climate_cfg(scenario = "stationary")
)

hazard_cfg
```

::: callout-note
## What the configuration parameters do
| Parameter | Role |
|---|---|
| `data_path` | Path to the IBTrACS NA CSV (or packaged subset). |
| `historical_start_year` | First IBTrACS year used to fit storm rates. 1970 marks the start of reliable satellite-era observations in the Atlantic. |
| `search_radius_km` | Track-point gate radius per site. Keep at 800 km unless you have a specific reason to deviate; a storm 700 km away can still produce TS-strength winds at the site through its extended wind field. |
| `simulation_years` | Number of synthetic years to generate. |
| `climate` | A `climate_cfg` object. `make_climate_cfg(scenario = "stationary")` switches off all SST forcing. |
| `preset` / `advanced` | Hooks for tuning class thresholds, wind-radii caps, and the lambda-scaling mode. Defaults follow WMO conventions (TS ≥ 34 kt, HUR ≥ 64 kt). |
| `background_wind` | Optional `background_wind_cfg` object (see @sec-background-wind). Fills non-storm days with trade-wind speeds sampled from site-specific Weibull distributions. `NULL` (default) sets all non-storm days to 0 kt. |

See `?make_hazard_cfg` for the full parameter list.
:::

## Stages 1 & 2 — Calibration and Simulation {#sec-run}

`run_hazard_model()` executes the historical calibration (Stage 1) and stochastic simulation (Stage 2) in one call.
With a stationary `climate` config, no SST forcing is applied.

```{r}
#| label: run-baseline
out_base <- run_hazard_model(
  cfg     = hazard_cfg,
  targets = targets,
  seed    = SEED,
  verbose = FALSE
)
```

### What happens inside `run_hazard_model()`

**Step 1 — Read and clean IBTrACS** (`read_ibtracs_clean()`). Standardise columns, retain the USA best-track wind/pressure fields, convert wind radii to km, and filter to the North Atlantic.

**Step 2 — Gate track points.** Compute great-circle distance from every 6-hour track point to each site and keep only points within `search_radius_km`.

**Step 3 — Compute site-level winds** (`compute_site_winds_full()`). Determine the site's quadrant relative to the storm centre, pull the matching directional wind radii (R34 / R50 / R64), estimate RMW (`estimate_RMW_knaff()`), evaluate the parametric wind profile, and apply a forward-motion asymmetry adjustment. Output: `site_wind_kt` per track point.

**Step 4 — Aggregate into events** (`make_storm_events()`). Group track points by storm and summarise into one event per (storm × site) with peak site wind, intensity and pressure summaries, mean RMW, timing, and the gated-point count. Classify by peak site wind into TS (34–63 kt) or HUR (≥ 64 kt).

**Step 5 — Estimate annual rates** (`compute_annual_counts()` → `compute_lambda_table()`). Convert per-class annual counts into mean rates `lambda` and estimate the Gamma overdispersion `k_hat` that controls year-to-year clustering.

**Step 6 — Simulate annual counts** (`simulate_twolevel_counts()`). For `simulation_years` synthetic years, draw the activity multiplier `A` from `Gamma(k_hat, k_hat)`, then total impactful storm counts from a Poisson mean conditional on `A`, then split into TS vs HUR via a binomial draw at `p_hurricane_base` (in stationary mode).

::: callout-tip
## Inspect the calibrated rates first
The first thing to look at after a run is `out_base$rates`. If a site's `lambda` is far from a published HURDAT2 rate, the rest of the workflow will inherit that bias — fix it before downstream analysis. Tutorial 2's Tier 2 check formalises this.
:::

```{r}
#| label: inspect-rates-quick
out_base$rates |>
  select(location, storm_class, lambda) |>
  mutate(lambda = round(lambda, 3)) |>
  arrange(location, storm_class)
```

### Inspecting `out_base` {#sec-outputs}

The return value is a named list with one tidy table per stage.

#### Rates

```{r}
#| label: inspect-rates
out_base$rates
```

| Column | Description |
|---|---|
| `location` | Target site. |
| `storm_class` | `TS` (34–63 kt site wind) or `HUR` (≥ 64 kt). |
| `lambda` | Mean annual rate of class events. |
| `n_years` | Historical years contributing to the estimate. |
| `prob_annual` | Probability of ≥ 1 event per year, $1 - e^{-\lambda}$. |
| `prob_none` | Probability of zero events, $e^{-\lambda}$. |

A `lambda` of 1.2 for TS means roughly 1.2 tropical-storm impacts per year on average; the actual count varies because of the Gamma-distributed activity multiplier.

#### Fit

```{r}
#| label: inspect-fit
out_base$fit
```

| Column | Description |
|---|---|
| `location` | Target site. |
| `k_hat` | Gamma-Poisson overdispersion. Smaller values (~ 2–5) imply burstier seasons; larger values approach pure Poisson. |
| `mean_annual_total` / `var_annual_total` | First two moments of the total annual count. |
| `beta_sst` | SST rate-scaling coefficient (0 in stationary mode). |
| `gamma_intensity` | Intensity / class-share shift coefficient (0 in stationary mode). |
| `p_hurricane_base` | Baseline probability that an impactful event is HUR-class. |

#### Sim

```{r}
#| label: inspect-sim
head(out_base$sim)
```

| Column | Description |
|---|---|
| `location` | Target site. |
| `sim_year` | Synthetic year index, 1 to `N_SIM`. |
| `n_total` / `n_tc` | Total impactful storms in the year. |
| `n_ts` | TS-class events. |
| `n_hur` | HUR-class events. |
| `A` | Activity multiplier from `Gamma(k_hat, k_hat)`; > 1 = hyperactive season, < 1 = quiet. |
| `p_hur` | Hurricane share applied (constant in stationary mode). |

This table drives the daily downscaling and all downstream return-period and exceedance work.

#### Events

```{r}
#| label: inspect-events
head(out_base$events)
```

One row per (historical storm × site) with site-attenuated wind:

| Column | Description |
|---|---|
| `storm_id` | IBTrACS SID (e.g. `2017242N16333` for Irma). |
| `year` | Season year. |
| `peak_wind_kt` | **Peak wind at the site** — the model-computed value, accounting for distance and wind-field structure. |
| `storm_intensity_kt` | The storm's lifetime peak intensity, regardless of location. |
| `min_pressure_hpa` / `pressure_deficit_hpa` | Central pressure and maximum environment-minus-centre deficit. |
| `rmw_mean_km` | Mean radius of maximum wind. |
| `storm_class` | Site classification (`TS` or `HUR`). |
| `location` | Target site. |

The distinction between `peak_wind_kt` and `storm_intensity_kt` matters: a Cat-5 storm (≈ 155 kt) passing 400 km from a site might only produce Cat-1 winds (~ 70 kt) there. The Holland profile captures that attenuation.

#### Trackpoints

```{r}
#| label: inspect-trackpoints
head(out_base$trackpoints[["Saba"]])
```

The lowest-level output: every gated 6-hourly IBTrACS track point per site, with all computed wind-field columns (`site_wind_kt`, `dist_km`, `bearing_deg`, `RMW_km`, etc.). Useful for diagnostics and spot-checking individual storms.

## Stage 3 — Daily Wind Time Series {#sec-daily}

`generate_daily_hazard_impact_spatial()` converts the per-year storm counts into a daily site-level wind series for every requested simulated year and location.

```{r}
#| label: daily-multisite
daily_by_loc <- generate_daily_hazard_impact_spatial(
  out         = out_base,
  location    = targets$name,
  sim_years   = seq_len(N_SIM),
  year0       = YEAR0,
  gust_factor = GUST_FACTOR,
  damage      = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario    = "stationary",
  seed        = SEED
)
```

The returned object is a named list of daily tibbles — one per location.

### What the function does

**Step 1 — Build event libraries.** For each location, the model builds a stratified empirical library of historical events with their physical attributes paired (peak wind, duration, pressure, RMW). A shared basin-level event pool is constructed across locations to keep multi-site sampling consistent.

**Step 2 — Sample events for each simulated year.** For each `sim_year`, draw the prescribed number of TS and HUR events from the library. Attribute pairing is preserved so wind, duration, pressure, and RMW remain physically consistent.

**Step 3 — Place events on the calendar.** Each sampled event gets a start date drawn from the empirical day-of-year distribution, concentrating events in the August–October peak.

**Step 4 — Spread each event into a daily wind pulse.** The `cosine` pulse shape gives a smooth bell-shaped envelope rising to the event peak and decaying over the event duration. `pulse_shape = "uniform"` is also available for a flat box.

**Step 5 — Combine overlapping events.** Multiple events that overlap in calendar time are reduced to the daily maximum sustained wind. This avoids double-counting while still capturing compound seasons.

**Step 6 — Derived daily fields.** Apply the gust factor, the damage function (`damage = list(method = "intensity")` returns the normalized intensity-based damage proxy; `"powerlaw"` is also available), and forward dominant-event metadata (`event_id`, `event_class`, `pressure_hpa`, `rmw_km`).

::: callout-tip
## Common-random-numbers comparisons
Pass the same `seed` and the same `sim_years` vector across baseline and scenario runs, and any difference in daily output isolates the climate forcing. This is the pattern used in Tutorial 3 and the stress-test experiment script.
:::

::: callout-note
## Pinned events
`pinned_sids` lets you force a specific historical SID into one or more sampled event slots — used by the stress-test workflow in Tutorial 4 to keep a focal storm (e.g. Irma) present across every scenario year.
:::

### Daily output schema

```{r}
#| label: daily-saba-head
head(daily_by_loc[["Saba"]])
```

| Column | Description |
|---|---|
| `location` | Target site. |
| `sim_year` | Simulated year index. |
| `scenario` | Scenario label carried through to downstream tables. |
| `date` | Calendar date (year derived from `year0`). |
| `wind_kt` | Daily site-level sustained wind (kt). |
| `wind_gust_kt` | `wind_kt × gust_factor`. |
| `surge_m` | Estimated storm surge (m), derived from event properties. |
| `event_id` / `event_class` | Identifier and class (`TS` / `HUR`) of the dominant event on that day. |
| `pressure_hpa` / `pressure_deficit_hpa` | Central pressure and deficit of the dominant event. |
| `rmw_km` | Radius of maximum wind of the dominant event. |
| `damage_intensity` | Normalised hazard intensity on a 0–1 scale. |
| `damage_rate` | Daily fractional damage rate from the damage function. |
| `cum_damage` | Cumulative damage within each simulated year. |

### Background Wind {#sec-background-wind}

By default, every day without a TC event has `wind_kt = 0`.
This is intentional for hazard-return-period work (zeros represent TC-free days cleanly), but it makes the daily series unsuitable for applications that need a realistic non-storm wind envelope — energy yield modelling, chronic-exposure assessments, port disruption studies.

`make_background_wind_cfg()` specifies Weibull marginal distributions for the trade-wind background, with an optional Gaussian copula for spatial correlation across sites and an AR(1) coefficient for day-to-day persistence.
On each simulated day, a background wind speed is drawn from the site- and month-specific Weibull distribution; the storm-day wind from the TC pulse then replaces it via `pmax()`, so the TC signal always dominates on active days.

#### Configuring background wind

```{r}
#| label: bg-wind-cfg
# Helper: constant Weibull parameters across all months for one location.
# In production, fit shape/scale separately for each month from observed
# non-TC wind records (e.g. airport anemometer or reanalysis data).
make_months <- function(shape, scale) {
  tibble::tibble(month = 1:12, shape = shape, scale = scale)
}

# Representative Dutch Caribbean trade-wind climatology.
# shape = 2.0 (Rayleigh-like), scale = 13 kt -> mean ~ 11.5 kt year-round.
bg_cfg <- make_background_wind_cfg(
  weibull_params = list(
    St_Martin = make_months(shape = 2.0, scale = 13),
    Saba      = make_months(shape = 2.0, scale = 13),
    Statia    = make_months(shape = 2.0, scale = 13)
  ),
  ar1 = 0.4   # moderate day-to-day persistence
)
```

The `ar1` coefficient controls how strongly today's background wind predicts tomorrow's in standardised normal space.
`ar1 = 0` gives independent daily draws; `ar1 = 0.4` introduces realistic multi-day persistence similar to synoptic trade-wind patterns.
For multi-site runs, `cor_matrix` lets you specify site-to-site Pearson correlations so that calm or windy spells are spatially coherent.

#### Embedding in the hazard configuration (recommended)

The cleanest production pattern attaches the background wind configuration to `make_hazard_cfg()`.
It is then stored in `out$cfg` and picked up automatically by `generate_daily_hazard_impact_spatial()` without any extra argument:

```{r}
#| label: hazard-cfg-bg
#| eval: false
hazard_cfg_bg <- make_hazard_cfg(
  data_path             = DATA_PATH,
  search_radius_km      = SEARCH_RADIUS_KM,
  historical_start_year = HIST_START_YEAR,
  simulation_years      = N_SIM,
  climate               = make_climate_cfg(scenario = "stationary"),
  background_wind       = bg_cfg          # <-- attach here once
)

out_bg <- run_hazard_model(
  cfg     = hazard_cfg_bg,
  targets = targets,
  seed    = SEED,
  verbose = FALSE
)

# background_wind is picked up from out_bg$cfg automatically
daily_bg_list <- generate_daily_hazard_impact_spatial(
  out         = out_bg,
  location    = targets$name,
  sim_years   = seq_len(N_SIM),
  year0       = YEAR0,
  gust_factor = GUST_FACTOR,
  damage      = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario    = "stationary",
  seed        = SEED
)
```

#### Passing background wind directly (exploratory use)

If you already have a model output object and want to try different background configurations without re-running the full calibration, pass `background_wind` directly to `generate_daily_hazard_impact_spatial()`.
An explicit argument always overrides anything stored in `out$cfg`:

```{r}
#| label: daily-bg-direct
daily_bg_list <- generate_daily_hazard_impact_spatial(
  out             = out_base,
  location        = targets$name,
  sim_years       = 1:50,
  year0           = YEAR0,
  gust_factor     = GUST_FACTOR,
  damage          = list(method = "intensity"),
  pulse_shape     = "cosine",
  scenario        = "stationary",
  background_wind = bg_cfg,
  seed            = SEED
)
```

#### Effect on the daily series

Comparing 50 simulated years for Saba shows the reduction in zero-wind days and the resulting uplift in non-storm-period mean wind:

```{r}
#| label: bg-wind-compare
bind_rows(
  daily_by_loc[["Saba"]] |>
    dplyr::filter(sim_year <= 50) |>
    dplyr::mutate(run = "No background"),
  daily_bg_list[["Saba"]] |>
    dplyr::mutate(run = "With background")
) |>
  dplyr::group_by(run) |>
  dplyr::summarise(
    pct_zero_wind      = round(mean(wind_kt == 0) * 100, 1),
    mean_wind_all_kt   = round(mean(wind_kt), 2),
    mean_wind_nonzero  = round(mean(wind_kt[wind_kt > 0]), 2),
    .groups = "drop"
  )
```

`pct_zero_wind` drops to near zero once background wind is active.
`mean_wind_all_kt` rises to reflect the trade-wind baseline; `mean_wind_nonzero` stays close to the TC-period mean because those days are dominated by the storm pulse, not the background.

::: callout-note
## TC-day winds are unaffected
Background wind is applied via `pmax(bg, tc_pulse)`.
On days when the TC pulse exceeds the background draw, the TC value is kept unchanged.
Return-level statistics for TS and HUR thresholds are therefore unaffected by the background wind setting.
:::

## Annual Activity Summary {#sec-annual}

Collapse the daily series into per-site annual statistics.

```{r}
#| label: annual-summary
daily_all <- bind_rows(daily_by_loc)

annual_summary <- daily_all |>
  group_by(location, sim_year) |>
  summarise(
    peak_wind_kt  = max(wind_kt,    na.rm = TRUE),
    n_ts_days     = sum(wind_kt >= 34, na.rm = TRUE),
    n_hur_days    = sum(wind_kt >= 64, na.rm = TRUE),
    annual_damage = sum(damage_rate, na.rm = TRUE),
    .groups = "drop"
  )

activity_means <- annual_summary |>
  group_by(location) |>
  summarise(
    mean_peak_wind_kt  = mean(peak_wind_kt),
    mean_ts_days       = mean(n_ts_days),
    mean_hur_days      = mean(n_hur_days),
    mean_annual_damage = mean(annual_damage),
    p_hur_year         = mean(n_hur_days > 0),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

activity_means
```

`p_hur_year` is the model-implied probability of at least one hurricane day in a season — a useful quick check that the simulation reproduces the expected basin behaviour.

## Key Parameters — What to Tune and Why {#sec-tuning}

- **`simulation_years`.** Controls Monte Carlo noise. Use ≥ 1,000; 2,000+ for return-period work above ~ 50 yr.
- **`search_radius_km`.** Track-point gate radius per site. 800 km is conservative and usually captures all relevant storms. Reducing it can starve the rate fit; increasing it pulls in tracks that contribute negligible site wind.
- **`historical_start_year`.** Start of the calibration window. 1970 is the practical minimum for reliable Atlantic best tracks; later starts shorten the record and inflate sampling uncertainty in `lambda` and `k_hat`.
- **`gust_factor`.** Sustained-to-gust conversion. Use 1.0 when comparing model output to IBTrACS sustained wind; ~1.3 when the downstream model wants gust forcing.
- **`damage`.** `list(method = "intensity")` returns a normalised wind-intensity damage proxy; `"powerlaw"` exposes a configurable wind-cubed-style damage curve.
- **`pulse_shape`.** `"cosine"` gives a smooth bell envelope; `"uniform"` is a flat-top alternative useful for sensitivity testing.
- **`background_wind`.** Pass a `make_background_wind_cfg()` object to fill non-storm days with realistic trade-wind speeds. Keep `NULL` for return-level work (zeros represent TC-free days unambiguously); use a non-`NULL` value for energy yield, chronic exposure, or port disruption analyses that need a continuous wind record. See @sec-background-wind for the full workflow.

## Where to Go Next {#sec-next}

* [Tutorial 2 — Model Validation](tutorial_2_validation.qmd) runs the three-tier validation suite (hindcast return levels, rate sanity, wind-field spot-checks) on this same `out_base`.
* [Tutorial 3 — Climate Change](tutorial_3_climate_change.qmd) replaces the stationary `climate_cfg` with KNMI'23 and IPCC AR6 scenarios and shows how to compare baseline vs scenario daily series with common random numbers.
* The stress-test workflow in `inst/extcode/05-stress-test-experiment.R` selects representative high-impact years from the baseline and replays them under climate scenarios with a pinned focal event.
