---
title: "Tutorial 4 — Climate Stress-Test Experiment"
subtitle: "Hurricane Wind Hazard Model for the Caribbean"
author: ""
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    smooth-scroll: true
    self-contained: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup-hidden
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

::: callout-note
## Prerequisites
This tutorial builds on [Tutorial 3](tutorial_3_climate_change.qmd).
You should already be familiar with `make_hazard_cfg()`, `run_hazard_model()`, and named KNMI'23 climate scenarios.
:::

This tutorial shows how to design a **climate stress-test experiment**: pick a handful of high-impact synthetic years from a long stationary baseline, then replay those *same* years under future-climate scenarios using common random numbers (CRN).
The result is a directly comparable set of daily wind traces where any difference between baseline and future is attributable to climate forcing alone.

## What This Tutorial Covers {#sec-what-this-covers}

By the end, you will be able to:

- generate a long stationary baseline and rate years on a composite stress score
- pick representative high-impact years at fixed severity quantiles (Q50–Q99)
- replay those year indices under KNMI'23 scenarios with CRN, pinning the dominant event
- visualize baseline-vs-scenario daily wind traces in a single faceted panel

```{r}
#| label: setup-packages
library(ipdcstorm)
library(dplyr)
library(ggplot2)
```

## Experiment Configuration {#sec-config}

A small block of switches controls the entire experiment.
The defaults below run quickly for the vignette; for a production stress-test, raise `n_sim` to 2,000 or more.

```{r}
#| label: config
seed            <- 42L
n_sim           <- 500L                       # raise to 2000+ in production
irma_wind_kt    <- 80                         # IRMA-like site-level threshold (kt)
quantile_probs  <- c(0.50, 0.80, 0.90, 0.95, 0.99)
knmi_scenarios  <- c("knmi_Ld", "knmi_Hd")    # low-dry and high-dry KNMI'23
target_year     <- 2100                       # 2050 (near-term) or 2100 (end-of-century)

# Composite rating weights (must sum to 1)
w_peak   <- 0.5   # severity (max site-level peak wind)
w_damage <- 0.5   # accumulation (integrated portfolio damage)

targets <- tibble::tribble(
  ~name,         ~lat,     ~lon,
  "St_Martin",   18.0708, -63.0501,
  "Saba",        17.6350, -63.2300,
  "Statia",      17.4890, -62.9740,
  "Puerto_Rico", 18.4655, -66.1057,
  "Miami",       25.7617, -80.1918
)

ibtracs_demo <- system.file("extdata", "ibtracs_demo.csv", package = "ipdcstorm")
```

## Part 1 — Baseline & Year Selection {#sec-baseline}

We first run a stationary baseline of `n_sim` synthetic years and downscale every year to daily wind and damage at every target location.

```{r}
#| label: run-baseline
cfg_base <- make_hazard_cfg(
  data_path        = ibtracs_demo,
  simulation_years = n_sim,
  climate          = make_climate_cfg(scenario = "stationary")
)

out_base <- run_hazard_model(cfg_base, targets = targets, seed = seed, verbose = FALSE)

daily_base_list <- generate_daily_hazard_impact_spatial(
  out         = out_base,
  location    = targets$name,
  sim_years   = seq_len(n_sim),
  year0       = 2000L,
  gust_factor = 1.0,
  damage      = list(method = "intensity"),
  pulse_shape = "cosine",
  scenario    = "baseline",
  seed        = seed
)
```

### Annual metrics across all sites {#sec-annual-metrics}

Each year is summarised on basin-wide metrics: peak site-level wind, hurricane- and storm-day counts, and integrated damage across all sites.

```{r}
#| label: annual-metrics
daily_base <- bind_rows(daily_base_list)

annual_all <- daily_base |>
  group_by(sim_year, location) |>
  summarise(
    peak_wind_kt = max(wind_kt, na.rm = TRUE),
    n_hur_days   = sum(wind_kt >= 64, na.rm = TRUE),
    n_storm_days = sum(wind_kt >= 34, na.rm = TRUE),
    total_damage = sum(damage_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(sim_year) |>
  summarise(
    peak_wind_max    = max(peak_wind_kt),
    hur_days_total   = sum(n_hur_days),
    storm_days_total = sum(n_storm_days),
    damage_total     = sum(total_damage),
    .groups = "drop"
  )
```

### Composite stress rating {#sec-rating}

Candidate years must contain at least one IRMA-class event (peak wind ≥ `r irma_wind_kt` kt).
Surviving years are rank-normalised on two metrics and combined into a single rating in `[0, 1]`:

| Metric | Meaning |
|--------|---------|
| `r_peak`   | Worst single-site peak wind (severity; loss scales ~wind³). |
| `r_damage` | Sum of `damage_rate` across days and sites (integrated impact). |

```{r}
#| label: rate-candidates
irma_candidates <- annual_all |>
  filter(peak_wind_max >= irma_wind_kt)

n_cand <- nrow(irma_candidates)

irma_rated <- irma_candidates |>
  mutate(
    r_peak   = rank(peak_wind_max, ties.method = "average") / n_cand,
    r_damage = rank(damage_total,  ties.method = "average") / n_cand,
    rating   = w_peak * r_peak + w_damage * r_damage
  ) |>
  arrange(rating)

# Pick years at the target quantiles of the rating distribution
sel_pos      <- pmax(1L, ceiling(quantile_probs * n_cand))
selected_ids <- irma_rated$sim_year[sel_pos]

irma_rated |>
  filter(sim_year %in% selected_ids) |>
  mutate(
    quantile = paste0("Q", round(quantile_probs * 100), "%")
  ) |>
  select(quantile, sim_year, peak_wind_max, hur_days_total,
         damage_total, rating)
```

### Pin the focal event {#sec-focal}

For each selected year we identify the SID of the basin-level peak hurricane day.
Pinning this SID into every scenario run guarantees the focal event always appears, with its intensity and duration perturbed by the scenario's `delta_sst` — all other storm slots remain free to vary with climate.

```{r}
#| label: focal-sids
baseline_stress <- lapply(daily_base_list, function(df) {
  df[df$sim_year %in% selected_ids, ]
})

focal_sids <- setNames(vector("list", length(selected_ids)),
                       as.character(selected_ids))

for (yr in selected_ids) {
  yr_hur <- bind_rows(lapply(baseline_stress, function(df) {
    df[df$sim_year == yr & !is.na(df$event_class) & df$event_class == "HUR", ]
  }))
  if (nrow(yr_hur) == 0L) {
    focal_sids[[as.character(yr)]] <- NA_character_
    next
  }
  peak_row <- yr_hur[which.max(yr_hur$wind_kt), ]
  focal_sids[[as.character(yr)]] <- sub("_y[0-9]+_[0-9]+$", "",
                                        peak_row$event_id[1L])
}

focal_sids
```

## Part 2 — KNMI'23 Climate Replays {#sec-replays}

The same `seed` and the same `sim_years = 1:n_sim` vector are used for every scenario.
This is the **common random numbers** trick: the RNG is consumed at exactly the same rate in every run, so year `i` in the baseline corresponds to year `i` in the scenario.
Differences between traces therefore isolate climate forcing rather than sampling noise.

::: callout-tip
## Why iterate the full `1:n_sim`?
Passing only `selected_ids` to the scenario run would put its first iteration at the initial RNG state, while the baseline reached those same indices after consuming hundreds of prior draws.
The result would be completely different event sampling — unrelated to climate.
:::

```{r}
#| label: run-scenarios
cc_stress <- vector("list", length(knmi_scenarios))
names(cc_stress) <- knmi_scenarios

for (scen in knmi_scenarios) {
  cfg_cc <- make_hazard_cfg(
    data_path        = ibtracs_demo,
    simulation_years = n_sim,
    climate          = make_climate_cfg(scenario   = scen,
                                        target_year = target_year,
                                        perturb     = TRUE)
  )

  out_cc <- run_hazard_model(cfg_cc, targets = targets,
                             seed = seed, verbose = FALSE)

  cc_daily_all <- generate_daily_hazard_impact_spatial(
    out         = out_cc,
    location    = targets$name,
    sim_years   = seq_len(n_sim),
    year0       = 2000L,
    gust_factor = 1.0,
    damage      = list(method = "intensity"),
    pulse_shape = "cosine",
    scenario    = scen,
    seed        = seed,
    pinned_sids = focal_sids
  )

  cc_stress[[scen]] <- lapply(cc_daily_all, function(df) {
    df[df$sim_year %in% selected_ids, ]
  })
}
```

## Part 3 — Visualize the Stress-Test Traces {#sec-viz}

Each panel below is one selected year, with the stationary baseline (black) overlaid against the KNMI'23 scenarios (blue = low-dry, red = high-dry).
Reference dashes mark the tropical-storm (34 kt) and hurricane (64 kt) thresholds.

```{r}
#| label: assemble-plot-data
focal_location <- "Saba"

scenario_labels <- c(
  "Baseline" = "Baseline (stationary)",
  "knmi_Ld"  = "KNMI L-d (low warming, dry)",
  "knmi_Hd"  = "KNMI H-d (high warming, dry)"
)
scenario_colors <- c(
  "Baseline" = "black",
  "knmi_Ld"  = "#4575b4",
  "knmi_Hd"  = "#d73027"
)

# Hurricane season window (Jun 1 – Dec 31, non-leap)
season_doy_start <- 152L
season_doy_end   <- 365L
month_doy_all    <- cumsum(c(1L, 31L, 28L, 31L, 30L, 31L,
                             30L, 31L, 31L, 30L, 31L, 30L))
season_months    <- 6:12
month_doy_starts <- month_doy_all[season_months]
month_labels     <- month.abb[season_months]

base_df <- baseline_stress[[focal_location]] |>
  mutate(scenario = "Baseline",
         doy      = as.integer(format(date, "%j")))

knmi_df <- bind_rows(lapply(knmi_scenarios, function(scen) {
  cc_stress[[scen]][[focal_location]] |>
    mutate(scenario = scen,
           doy      = as.integer(format(date, "%j")))
}))

trace_labels <- setNames(
  paste0("Trace ", seq_along(selected_ids),
         "  \u2014  Q", round(quantile_probs * 100), "%"),
  as.character(selected_ids)
)

plot_df <- bind_rows(knmi_df, base_df) |>
  filter(doy >= season_doy_start, doy <= season_doy_end) |>
  mutate(
    scenario    = factor(scenario, levels = c(knmi_scenarios, "Baseline")),
    trace_label = factor(trace_labels[as.character(sim_year)],
                         levels = trace_labels)
  )
```

```{r}
#| label: plot-traces
#| fig-width: 9
#| fig-height: 11
ggplot(plot_df, aes(x = doy, y = wind_kt)) +
  geom_hline(yintercept = 34, linetype = "dashed",
             colour = "#E69F00", linewidth = 0.35, alpha = 0.8) +
  geom_hline(yintercept = 64, linetype = "dashed",
             colour = "#d73027", linewidth = 0.35, alpha = 0.8) +
  geom_line(data = ~ filter(.x, scenario != "Baseline"),
            aes(colour = scenario), linewidth = 0.9, alpha = 0.9) +
  geom_line(data = ~ filter(.x, scenario == "Baseline"),
            aes(colour = scenario), linewidth = 1.2) +
  scale_colour_manual(values = scenario_colors,
                      labels = scenario_labels,
                      breaks = c("Baseline", knmi_scenarios),
                      name   = NULL) +
  scale_x_continuous(limits = c(season_doy_start, season_doy_end),
                     breaks = month_doy_starts,
                     labels = month_labels,
                     expand = expansion(mult = c(0.01, 0.01))) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.08))) +
  facet_wrap(~ trace_label, nrow = 5, ncol = 1, scales = "fixed") +
  labs(
    x = NULL,
    y = "Sustained wind speed (kt)",
    title = sprintf("%s \u2014 Stress-test wind traces: baseline vs KNMI\u201923",
                    focal_location),
    subtitle = sprintf(
      "%d high-impact years from %d synthetic years (seed = %d, target year %d)",
      length(selected_ids), n_sim, seed, target_year),
    caption = "Dashed: TS (34 kt, orange) and hurricane (64 kt, red) thresholds"
  ) +
  theme_light(base_size = 10) +
  theme(
    legend.position  = "bottom",
    legend.key.width = unit(1.4, "cm"),
    strip.text       = element_text(size = 8.5, face = "bold"),
    panel.spacing.y  = unit(0.4, "lines"),
    plot.subtitle    = element_text(size = 8, colour = "grey40"),
    plot.caption     = element_text(size = 7, colour = "grey50", hjust = 0)
  ) +
  guides(colour = guide_legend(nrow = 1,
                               override.aes = list(linewidth = 1.1)))
```

## Where to Go Next {#sec-next}

- Switch `target_year` to `2050` to compare near-term vs end-of-century forcing.
- Add `knmi_Ln` / `knmi_Hn` to `knmi_scenarios` for the wet variants.
- Tune `w_peak` / `w_damage` to bias year selection toward severity vs accumulation.
- Change `focal_location` to view the same experiment from a different island.
- For a production run, point `data_path` at the full IBTrACS NA CSV and raise `n_sim` to ≥ 2,000.
